\documentclass{beamer}

\mode<presentation> {

%%%%%%%THEME%%CHOICE%%%%%%

%\usetheme{default}
%\usetheme{AnnArbor}
%\usetheme{Antibes}
%\usetheme{Bergen}
%\usetheme{Berkeley}
%\usetheme{Berlin}
%\usetheme{Boadilla}
%\usetheme{CambridgeUS}
%\usetheme{Copenhagen}
%\usetheme{Darmstadt}
%\usetheme{Dresden}
%\usetheme{Frankfurt}
%\usetheme{Goettingen}
%\usetheme{Hannover}
%\usetheme{Ilmenau}
%\usetheme{JuanLesPins}
%\usetheme{Luebeck}
\usetheme{Madrid}
%\usetheme{Malmoe}
%\usetheme{Marburg}
%\usetheme{Montpellier}
%\usetheme{PaloAlto}
%\usetheme{Pittsburgh}
%\usetheme{Rochester}
%\usetheme{Singapore}
%\usetheme{Szeged}
%\usetheme{Warsaw}

%%%%%COLOUR%%PALLETTE%%%%%

%\usecolortheme{albatross}
\usecolortheme{beaver}
%\usecolortheme{beetle}
%\usecolortheme{crane}
%\usecolortheme{dolphin}
%\usecolortheme{dove}
%\usecolortheme{fly}
%\usecolortheme{lily}
%\usecolortheme{orchid}
%\usecolortheme{rose}
%\usecolortheme{seagull}
%\usecolortheme{seahorse}
%\usecolortheme{whale}
%\usecolortheme{wolverine}

%%%%%REMOVE%%FOOTER%%%%%

%\setbeamertemplate{footline} 

%%%%%REPLACE%%FOOTER%%WITH%%SLIDE%%COUNT%%%%%

%\setbeamertemplate{footline}[page number] 

%%%%%REMOVE%%NAVI%%SYMBOLS%%%%%

%\setbeamertemplate{navigation symbols}{} 
}

\usepackage{graphicx} 
\usepackage{booktabs} 
\usepackage{multicol}
\usepackage{tikz}
\usepackage{bussproofs}
\setbeamercolor{block title}{bg=red!30,fg=black}
\setbeamertemplate{itemize item}{\color{red!35}$\blacksquare$}
\setbeamercolor{local structure}{fg=darkred}
%show TOCs highlighting current section at beginning of section
\AtBeginSection[]
{
    \begin{frame}
        \frametitle{Table of Contents}
        \tableofcontents[currentsection]
    \end{frame}
}
\newcommand{\du}{\Diamond_\uparrow}
\newcommand{\dl}{\Diamond_\leftarrow}
\newcommand{\bu}{\Box_\uparrow}
\newcommand{\bl}{\Box_\leftarrow}
\title[Axiomatic Potentialism]{Axiomatic Potentialism} 
% The short title appears at the bottom of every slide, 
% the full title is only on the title page

\author{Chris Scambler} 
\institute[ASC] 
% Your institution as it will appear on the bottom of every slide, 
% may be shorthand to save space
{
All Souls College, \\
Oxford University \\ 
\medskip
\textit{chris.scambler@all-souls.ox.ac.uk} 
}
\date{\today} 

\begin{document}

\begin{frame}
\titlepage 
\end{frame}

\begin{frame}
\frametitle{Overview} 
\tableofcontents 
\end{frame}
\section{Background}

\begin{frame}
\frametitle{Potentialism}
\onslide<2->
\begin{block}
    {\bf Potentialism} is the idea that a mathematical object (e.g. a set) is the sort 
    of thing that may \emph{merely possibly} exist.
\end{block}
\begin{itemize}
    \item<3-> E.g. a geometric object as a figure one can construct
    \item<4-> A set as a certain sort of data structure one could assemble
    \item<5-> Or perhaps a structure that is instantiated given enough objects.
    \item<6-> Ideas like this have deep roots in set theory, e.g. Zermelo and even Cantor
    \item<7-> Still deeper roots in mathematics in general.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Potentialism}
    \begin{itemize}
        \item The recent literature has seen two branches of study here:
        \begin{enumerate}
            \item<2->   Model-theoretic: study Kripke models whose worlds are 
                        structures with the accessibility relation (some refinement of)
                        the substructure relation.
            \item<3->   Axiomatic: Develop axiom systems designed to characterize 
                        this or that form of potentialism directly, without appeal to models.
        \end{enumerate}
        \item<4->   In each case interesting questions arise concerning the relation 
                    between assertions in the modal framework and in first order set theory.
        \item<5->   The model-theoretic side has seen thorough examination in the recent literature
        \item<6->   Here we will be focused on axiomatic potentialism, and on 
                    relations between potentialist axiom systems and their first order counterparts.
    \end{itemize}
    
\end{frame}

\section{Warm Up: Height Potentialism}

\begin{frame}
    \frametitle{Motivation}
    \begin{itemize}
        \item<2->   Imagine one has the ability to take things and 
                    make a set containing them.
        \item<3->   Imagine one is able to do this arbitrarily many times.
        \item<4->   Axiomatize this conception and work out its non-modal counterpart.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Language}
    \begin{block}{The Language $\mathcal{L}_0$}
        \begin{itemize}
            \item object variables $x, y, z$
            \item<2-> plural variables $X, Y, Z$
            \item<3-> $\wedge, \neg, \forall, =$
            \item<4-> $\Box$
            \item<5-> $\in$
        \end{itemize} 
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Axioms for the theory $\mathsf{L}$}
    \begin{block}{Logical Axioms}
    \begin{enumerate}
        \item<2-> Free FO logic
        \item<3-> $\mathsf{S4.2}$ modal logic + CBF (or free $\mathsf{S5}$)
        \item<4-> $\forall z[Xz \leftrightarrow Yz] \rightarrow X = Y$, rigidity,
        Choice, Comp
    \end{enumerate}
    \end{block}
    \onslide<5->
    \begin{block}{Set-theoretic axioms}
        \begin{enumerate}
        \item<6-> Extensionality, $\in$-rigidity, foundation
        \item<7-> $\Box \forall X \Diamond \exists x [Set(x, X)]$
        \item<8-> $\Diamond \exists X \Box \forall x[Xx \leftrightarrow \mathbb{N}x]$
        \item<9-> $\Diamond \exists X \Box \forall x[Xx \leftrightarrow x \subseteq y]$
        \item<10-> A modal translation of replacement
        \end{enumerate}
    \end{block}
\end{frame}
\begin{frame}
    \frametitle{Inconsistency?}
    Standard modal model theory validates the rule
    \begin{prooftree}
        \AxiomC{$\varphi \rightarrow \Box \psi$}
        \UnaryInfC{$\varphi \rightarrow \Box \forall x \psi$}
    \end{prooftree}
\onslide<2->
    \begin{equation}
        (Xx \leftrightarrow x \not \in x) \rightarrow \forall y \Box \neg Set(y, X)
    \end{equation}
\onslide<3->
    \begin{equation}
        (Xx \leftrightarrow x \not \in x) \rightarrow \Box \neg Set(y, X)
    \end{equation}
\onslide<4->
    \begin{equation}
        (Xx \leftrightarrow x \not \in x) \rightarrow \Box \forall y \neg Set(y, X)
    \end{equation}
\onslide<5->
 Hence the need for free logic.

\onslide<6-> UI = $\forall x [\forall y \varphi y \rightarrow \varphi x]$; 

\onslide<7-> Instantiation requires assumption of existence.
\end{frame}
\begin{frame}
    \frametitle{Relative Consistency}
\begin{itemize}
    \item In fact ZFC interprets $\mathsf{L}$.
\end{itemize}
\onslide<2->
\begin{block}{$t : \mathcal{L}_0 \times V \to \mathcal{L}_\in, (\varphi, T) \mapsto \psi(T)$}
    \begin{itemize}
        \item<3-> assign singular/plural variables even/odd numbered variables $t(x), t(X)$. (!)
        \item<4-> membership claims = id, commutes with propositional connectives
        \item<5-> $t(Xx)(T) := t(x) \in t(X) \subseteq T$
        \item<6-> $t(\forall x \varphi)(T) := \forall x \in T [t(\varphi)(T)]$
        \item<7-> $t(\forall X \varphi)(T) := \forall x \subseteq T [t(\varphi)(T)]$
        \item<8-> $t(\Box \varphi)(T) := \forall S \supseteq T [t(\varphi)(S)]$
    \end{itemize}
\end{block}
\onslide<9->
\begin{block}{Theorem}
  $\mathsf{L} \vdash \varphi$ implies $ZFC \vdash t(\varphi)(T)]$  
\end{block}
\onslide<10-> Note: convention on variables $T, S$.
\end{frame}
\begin{frame}
    \frametitle{Converse Translation}
\onslide<2->
\begin{block}{Mirroring theorem}
\onslide<3-> For $\varphi$ in $\mathcal{L}_\in$, let $\varphi^\Diamond$ be the result of 
prefixing all atomic formulas and existential quantifiers by $\Diamond$
(and universal quantifiers by $\Box$.) Then we have 
\onslide<4->
\[
    \Gamma \vdash_{FOL} \varphi 
    \Leftrightarrow 
    \Gamma^\Diamond \vdash_{\mathsf{L}} \varphi^\Diamond
\]
\end{block}
\onslide<5-> Note on replacement$^\Diamond$.
\onslide<6->
\begin{block}{Linnebo Interpretation Theorem}
  $\mathsf{L} \vdash ZFC^\Diamond$.  
\end{block}
\onslide<7-> Proof: use mirroring.
\end{frame}
\begin{frame}
    \frametitle{Axioms for the theory $\mathsf{L}$}
    \begin{block}{Logical Axioms}
    \begin{enumerate}
        \item Free FO logic
        \item $\mathsf{S4.2}$ modal logic + CBF
        \item $\forall z[Xz \leftrightarrow Yz] \rightarrow X = Y$, rigidity,
        Choice, Comp
    \end{enumerate}
    \end{block}
    \begin{block}{Set-theoretic axioms}
        \begin{enumerate}
        \item Extensionality, $\in$-rigidity, foundation
        \item $\Box \forall X \Diamond \exists x [Set(x, X)]$
        \item $\Diamond \exists X \Box \forall x[Xx \leftrightarrow \mathbb{N}x]$
        \item $\Diamond \exists X \Box \forall x[Xx \leftrightarrow x \subseteq y]$
        \item A modal translation of replacement
        \end{enumerate}
    \end{block}
\end{frame}

\begin{frame}
    \frametitle{Mini-conclusion}
\onslide<2->
\begin{block}{Equivalence}
    We have an exact proof-theoretic equivalence, $\mathsf{L} \equiv ZFC$.
\end{block}
\onslide<3->
\begin{block}{Bi-interpretation}
    In fact, ZFC $\vdash t(\phi^\Diamond)(T) \leftrightarrow \phi$;
    \onslide<4->and \\ L $\vdash Univ(T) \rightarrow (t(\phi)(T)^\Diamond \leftrightarrow \phi)$
\end{block}
\onslide<5->
This is essentially what Button calls `near-synonymy'.
\end{frame}

\section{Height and Width Potentialism}
\begin{frame}
    \frametitle{Motivation}
    \begin{itemize}
        \item<2->   Imagine one has the ability to take things and 
                    make a set containing them.
        \item<3->   Imagine one is also able to take a partial order and add a filter 
                    meeting all its (current) dense sets; 
        \item<4->   Or, equivalently, to take some things and add an enumerating function.
        \item<5->   Axiomatize this conception and work out its non-modal counterpart.
    \end{itemize}
\end{frame}
\begin{frame}
    \frametitle{Language}
    \begin{block}{The Language $\mathcal{L}_1$}
        \begin{itemize}
            \item object variables $x, y, z$
            \item<2-> plural variables $X, Y, Z$
            \item<3-> $\wedge, \neg, \forall, =$
            \item<4-> $\bu$, $\bl$, $\Box$
            \item<5-> $\in$
        \end{itemize} 
    \end{block}
\end{frame}
\begin{frame}
    \frametitle{Axioms for the theory $\mathsf{M}$}
    \begin{block}{Logical Axioms}
    \begin{enumerate}
        \item<2-> Free FO logic
        \item<3-> $\mathsf{S4.2}$ modal logic + CBF for each modal (or free $\mathsf{S5}$)
        \item<4-> $\Box \varphi \rightarrow \bu \varphi$, same for $\bl$.
        \item<5-> $\forall z[Xz \leftrightarrow Yz] \rightarrow X = Y$, rigidity,
        Choice, Comp
    \end{enumerate}
    \end{block}
    \onslide<6->
    \begin{block}{Set-theoretic axioms}
        \begin{enumerate}
        \item<7-> Extensionality, $\in$-rigidity, foundation
        \item<8-> $\Box \forall X \du \exists x [Set(x, X)]$
        \item<9-> $\du \exists X \Box \forall x[Xx \leftrightarrow \mathbb{N}x]$
        \item<10-> $\du \exists X \bu \forall x[Xx \leftrightarrow x \subseteq y]$
        \item<11-> $\Box \forall X \dl \exists f [f : \mathbb{N} \twoheadrightarrow X]$
        \item<12-> A modal translation of replacement
        \end{enumerate}
    \end{block}
\end{frame}
\begin{frame}{Some basic facts}
    \begin{itemize}
        \item<2-> $\mathsf{M}$ interprets ZFC under the translation 
                    $\varphi \mapsto \varphi^{\Diamond_\uparrow}$. 
        \item<3-> $\mathsf{M}$ interprets ZFC$^-$ under the translation 
                    $\varphi \mapsto \varphi^\Diamond$.
        \item<4-> $\mathsf{M}$ proves $\neg Pow^\Diamond$.
        \item<5-> $\mathsf{M}$ proves $V = HC^\Diamond$ and hence $SOA^\Diamond$.
        \item<6-> $\mathsf{M}$ proves the universal possibility of forcing.
        \item<7-> e.g., it proves it is possible for the continuum 
                    to exist and have a cardinality at least as great as 
                    any $\aleph$ number whose existence is provable 
                    in ZFC.
    \end{itemize}
\end{frame}
\begin{frame}
    \frametitle{Inconsistency?}
The axioms imply 
\[
    \Diamond \exists x (\Diamond_\leftarrow \exists y[y \subseteq x \wedge y = z] \wedge \Box_\uparrow \neg \exists y[y = z])
\]
\onslide<2-> 
abbreviate the formula in parentheses by $\Psi(x, z)$. 

\onslide<3-> 
By comprehension,
\[
    \Diamond \exists x \Psi(x, z) \wedge \exists X \forall y[Xy \leftrightarrow y \in z]
\]

\onslide<4-> 
By height potentialism/rigidty,
\[
    \Diamond \exists x \Psi(x, z) \wedge \du \exists w \forall y[y \in w \leftrightarrow y \in z]
\]
\onslide<5-> But then the rigidity/extensionality imply $w = z$ after all, 
so we have a contradiction.
\end{frame}
\begin{frame}
    \frametitle{Resolution}
The argument just sketched uses comprehension with arbitrary parameters:
\[\exists X\forall y[Xy \leftrightarrow y \in z]\]
\onslide<2-> And in the crucial application, it applies when we have no 
\emph{a priori} guarantee $z$ even exists (indeed this is what we are trying 
to establish.)
\onslide<3-> Natural solution: `restrict' comp to closed form:
\[
    \Box \forall z \Box \forall Z \exists X \forall y[Xy \leftrightarrow \varphi(y, z, Z)]
\]
\onslide<4-> Amounts to `restricting' ourselves to parameters that exist at 
the world of evaluation.
\end{frame}

\begin{frame}
    \frametitle{Relative Consistency}
\begin{block}{Key Point}
    \onslide<2-> It turns out that $\mathsf{M}$ exhibits roughly the same 
    relation to second order arithmetic together with the $\mathbf{\Pi}_1^1$
    -\emph{Perfect Set Property} as $\mathsf{L}$ to ZFC.
\end{block}
\begin{itemize}
    \item<3-> A set of reals is \emph{perfect} iff it is closed and has 
            no isolated points. There are only $2^{\aleph_0}$ such sets.
    \item<4-> A set of reals is said to \emph{have the perfect set property} iff it is 
            countable, or has a perfect subset.
    \item<5-> E.g. every closed set has the PSP.
    \item<6-> If the reals can be well-ordered (AC), not all sets have the PSP.
    \item<7-> The $\Gamma$-PSP is the assertion that every set of reals defined by a 
            formula of type $\Gamma$ has the PSP.
    \item<8-> The $\Gamma$-PSP implies no formula in $\Gamma$ defines a well-order of 
                the reals.
    \item<9-> $\Gamma$-definable sets are thus `regular' or `well behaved'.
\end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Relative Consistency}
\begin{block}{Intuitive idea}
    \onslide<2-> (From now on, I will ignore the difference between SOA 
    and ZFC$^-$+ V = HC. Replacement is formulated as collection.)
    \begin{itemize}
    \item<3-> We will use the fact that $\mathsf{T}$ = SOA + $\Pi_1^1$-PSP 
    $\equiv$ ZFC, and in fact $\mathsf{T}$ proves that $L[r]$ is a 
    model of ZFC for every real $r$. 
    \item<4-> 
    Our translation will be doubly parameterized, once by a real and 
    once by a transitive set.
    \item<5->
    Our interpretation for $\du$ will involve holding $r$ fixed and 
    climbing transitive sets in $L[r]$;
    \item<6-> 
    while our interpretation for $\dl$ will involve allowing new reals 
    to be added but not extending the height of the transitive set parameter.
    \end{itemize}
\end{block}
\end{frame}
\begin{frame}
    \frametitle{Relative consistency}
    \onslide<2-> Let $M \vDash SOA + \Pi_1^1 PSP$.
    \onslide<3->
    \begin{block}{$t : \mathcal{L}_1 \times M \times \mathbb{R}^M \to \mathcal{L}_\in, (\varphi, T, r) \mapsto \psi(T, r)$}
        \begin{itemize}
            \item<4-> assign singular/plural variables even/odd numbered variables $t(x), t(X)$.
            \item<5-> Membership = id, commutes with propositional connectives 
            \item<6-> $t(Xx)(T, r) := t(x) \in t(X)$
            \item<7-> $t(\forall x \varphi)(T, r) := \forall x \in T [t(\varphi)(T, r)]$
            \item<8-> $t(\forall X \varphi)(T, r) := \forall x \subseteq T [x \in L[r] \rightarrow t(\varphi)(T, r)]$
            \item<9-> $t(\bu \varphi)(T, r) := \forall S \supseteq T [ S \in L[r] \rightarrow t(\varphi)(S, r)]$
            \item<10-> $t(\bl \varphi)(T, r) := \forall s[ r \in L[s] \rightarrow \forall S \in L[s][ T \subseteq S \wedge rank(S) = rank(T) \rightarrow t(\varphi)(S, s)]]$
        \end{itemize}
    \end{block}
    \onslide<11->
\begin{block}{Theorem}
  $\mathsf{M} \vdash \varphi$ implies $\mathsf{T} \vdash \forall r, T \in L[r][t(\varphi)(X, r)]$  
\end{block}
\end{frame}
\begin{frame}
    \frametitle{Converse Translation}
\onslide<2->
\begin{block}{Theorem}
$\mathsf{M} \vdash \Pi_1^1 PSP^\Diamond$
\end{block}
\onslide<3->
\begin{block}{Proof (sketch)}
    \onslide<4-> Solovay: If only countably many reals are constructible from any $r$, then the $\Pi_1^1$-PSP 
    holds. 

    \onslide<5-> By mirroring, suffices to establish that (only countably many reals are constructible from $r$)$^\Diamond$ holds.

    \onslide<6-> Since we have ZFC$^{\du}$, we get that (there is a set of all reals constructible from $r$)$^{\du}$.

    \onslide<7-> Some absoluteness lemmas imply (there is a set of all reals constructible from $r$)$^\Diamond$.

    \onslide<8-> Since (all sets are countable)$^\Diamond$, the result follows.

\end{block}
\end{frame}
\begin{frame}{Interpretations}
    \onslide<2->
    \begin{block}{Equivalence}
        We thus have an exact proof-theoretic equivalence, $\mathsf{M} \equiv SOA + \Pi_1^1 PSP$.

        \onslide<3-> The latter is in fact equiconsistent with ZFC. $\mathsf{M} \equiv \mathsf{L} \equiv ZFC$.
    \end{block}
    \onslide<4->
    \begin{block}{Bi-interpretation}
        As before, SOA + $\Pi_1^1 PSP$ $\vdash t(\phi^\Diamond)(X, r) \leftrightarrow \phi$; 
        \onslide<5-> We also have a partial converse: for the language without $\dl$,
        M $\vdash Univ(T) \rightarrow (t(\phi)(T)^\Diamond \leftrightarrow \phi)$
    \end{block}
\onslide<6-> Thus we have at least `near'-near synonymy in this case. 

\onslide<7-> The exact details of $\dl$ are incidental anyway; if we formulate the theory with just $\Diamond$ and $\du$, we will get near synonymy. 
\end{frame}

\begin{frame}{Concluding remarks}
        \begin{itemize}
        \item<2-> The `purely quantificational picture' corresponding to the height + width 
        potentialist view enshrined in $\mathsf{M}$ 
        seems to be second order arithmetic + $\Pi_1^1$ PSP.
        
        \item<3-> What happens if we strengthen the regularity properties? 
        \item<4-> You get more large cardinals in inner models.
        \item<5-> In the modal theory, this corresponds to the $\du$-possibility of large cardinals.
        \end{itemize}
\end{frame}

\begin{frame}{Concluding remarks}
    \begin{itemize}
    \item These results have led me to consider second order arithmetic + regularity properties 
            as a possible `foundation for mathematics'.
    \item<2-> Such theories are strong in terms of consistency strength, and can interpret substantial portions 
            of the traditional large cardinal hierarchy.
    \item<3-> But they are also ontologically parsimonious, in only admitting countable sets, and offer hope   of certain benefits in applied mathematics -- e.g. the possibility of universal lebesgue measurability.
    \item<4-> Traditionally, the stronger regularity properties are believed to hold only in inner models, while the large cardinals are in the `real' universe of sets.
    \item<5-> But I am not convinced that there are decisive reasons for taking one of these theories, rather than the other, to hold only in inner models. 
    \item<6-> The potentialist perspective enshrined in $M$, and the regularity perspective in $T$, provide two 
    tightly equivalent ways of fleshing out the alternative viewpoint.
    \end{itemize}
\end{frame}


\begin{frame}
\Huge{\centerline{Thanks!}}
\end{frame}

\end{document} 